---
title: "STA 199 Project (title TBD)"
subtitle: "Proposal"
author: "TEAM 2: E. Wang, T. Chen, V. Garbelotto, S. Sengupta, K. Dong"
date: "03-07-23"
format: html
editor: visual
---

```{r}
#| label: load-pkgs

library(tidyverse)
```

# Data 1 - Billionaires

## Introduction and data

-   Identify the source of the data.

    -   Billionaires CSV File from the CORGIS Dataset Project.

    -   By Ryan Whitcomb\
        Version 2.0.0, created 5-17-16\
        Key Tags: money, rich, wealthy, people, person, billionaire

    Website located on 03-07-23:<https://think.cs.vt.edu/corgis/csv/billionaires/>

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

    -   "Researchers have compiled a multi-decade database of the super-rich. Building off the Forbes World's Billionaires lists from 1996-2014, scholars at Peterson Institute for International Economics have added a couple dozen more variables about each billionaire - including whether they were self-made or inherited their wealth. (Roughly half of European billionaires and one-third of U.S. billionaires got a significant financial boost from family, the authors estimate.)" - Overview located on CORGIS data set original site.

    -   This publication was cited as the origin of the data:<http://www.iie.com/publications/interstitial.cfm?ResearchID=2917>

-   Write a brief description of the observations.

    This data set compiles a list of billionaires from around the globe from 1996 until 2016. It ranks them from most to least wealthy, it tells us when the data of each billionaire was collected, their age, their net worth, whether this money was inherited, and more.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

    We would like to explore the correlation between inheritance and current net worth. For this, we would ask... Does receiving an inheritance impact the total net worth of each billionaire? Does birth place/country of citizenship have an impact on inheritance or net worth? Does wealth type impact net worth? Ultimately, did inheritance "create" the majority of the wealthiest billionaires in the world?

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

    We are interested in researching the correlation between inheritance and the net worth of the billionaires listed in this data set. Specifically, we would like to investigate whether inheritance had a key role in creating the world's richest people. We believe this to be true and that the majority of billionaires had help from an inheritance to build their net worth.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

    Freund, Caroline, and Sarah Oliver. "The origins of the superrich: the billionaire characteristics database." Peterson Institute for International Economics Working Paper 16-1 (2016).

    This article directly addresses the origins of how billionaires get their wealth and what characteristics contributed to the rise of their wealth.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-billionaires-data

billionaires <- read_csv("data/billionaires.csv")
```

# Data 2 - Elections

## Introduction and data

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

    This data was collected from CNN's 2016 election center. Ryan Whitcomb collected the data from CNN's website in June 2016. The data contains information on how many county votes and the proportion of the county vote each candidate in the 2016 election received.

    Link to data source: https://think.cs.vt.edu/corgis/csv/election/

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

How does geographic location (Northeast, South, West, Midwest) influence what proportion of a county's vote Donald Trump and Hillary Clinton received in the 2016 Presidential Election?

For this research topic we will be examining the varying presidential preferences of voters in different geographic regions of the US. We hypothesize that voters in the the South and Midwest more strongly prefer Donald Trump while voters in the Northeast and West more strongly prefer Hillary Clinton.

Below are the variables we will consider in our analysis:

Region (nominal categorical): Mutate the Location.State Abbreviation variable to group states into four regions - Northeast, South, West, and Midwest

Vote Data.Donald Trump.Percent of Votes (continuous numeric)

Vote Data.Hillary Clinton.Percent of Votes (continuous numeric)

## Literature

-   Find one published credible article on the topic you are interested in researching.

    -   Feinberg, M., Tullett, A. M., Mensch, Z., Hart, W., & Gottlieb, S. (2017). The political reference point: How geography shapes political identity. *PLOS ONE*, *12*(2). https://doi.org/10.1371/journal.pone.0171497

-   Provide a one paragraph summary about the article.

    -   The article "The Political Reference Point: How Geography Shapes Political Identity" provides valuable insights into the complex relationship between geography and political identity, highlighting the importance of considering the influence of regions on individual's party affiliation and voting preference. It discusses how social and historical forces such as the Civil Right Movement could shape the political identity of regions significantly, and also geographical factors such as climate, terrain, and resource availability could potentially influence people's values and beliefs. For example, people who live in areas with harsh climates may prioritize self-sufficiency and individualism. Hence, the political "blueness" or "redness" of people's locations also influence their voting preferences, exemplified in the presidential elections.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

    -   Our team's research question aims to investigate how the relationship between geography and the voting preference exemplified in the 2016 Presidential Election, building on the hypothesis and the conclusion from the journal article. Specifically, we divide the regions to Northeast, South, West, Midwest, and investigate the influence of the regions on the voters' preference towards Donald Trump and Hilary Clinton.

## Glimpse of data

```{r}
#| label: load-election-data

election <- read_csv("data/election.csv")
```

# Data 3

## Introduction and data

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-3

```
