---
title: "Project title"
subtitle: "Proposal"
author: "TEAM 2: E. Wang, T. Chen, V. Garbelotto, S. Sengupta, K. Dong"
date: "03-07-23"
format: html
editor: visual
---

```{r}
#| label: load-pkgs

library(tidyverse)
```

# Data 1

## Introduction and data

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-1

# add code here
```

# Data 2 - Elections

## Introduction and data

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

    This data was collected from CNN's 2016 election center. Ryan Whitcomb collected the data from CNN's website in June 2016. The data contains information on how many county votes and the proportion of the county vote each candidate in the 2016 election received.

    Link to data source: https://think.cs.vt.edu/corgis/csv/election/

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

How does geographic location (Northeast, South, West, Midwest) influence what proportion of a county's vote Donald Trump and Hillary Clinton received in the 2016 Presidential Election?

For this research topic we will be examining the varying presidential preferences of voters in different geographic regions of the US. We hypothesize that voters in the the South and Midwest more strongly prefer Donald Trump while voters in the Northeast and West more strongly prefer Hillary Clinton.

Below are the variables we will consider in our analysis:

Region (nominal categorical): Mutate the Location.State Abbreviation variable to group states into four regions - Northeast, South, West, and Midwest

Vote Data.Donald Trump.Percent of Votes (continuous numeric)

Vote Data.Hillary Clinton.Percent of Votes (continuous numeric)

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-election-data

election <- read_csv("data/election.csv")
```

# Data 3

## Introduction and data

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-3

```
