---
title: "Billionaire Wealth"
subtitle: "Report"
format: html
editor: visual
author: "Emily W., Valentina G., Teresa C., Karen D., & Sneha S."
execute:
  echo: true
---

# Introduction and Data

To the average person, the modern billionaire is at once mysterious and completely scrutable. Few will ever come face to face with a billionaire, yet many can name the likes of Jeff Bezos and Bill Gates and a fact or two about each. While these billionaires sit on enormous wealth, 47% of the world lives on less than \$6.85 per day, and 84% lives on less than \$30 per day \[1\]. Moreover, the past two decades have seen a steep increase in global inequality as the world's billionaires have captured an increasing proportion of global wealth \[2\]. Our realities of poverty among wealth and stark inequality thus incite questions about how world leaders should engage with billionaire wealth - was this wealth justly acquired? What are the ethics of wealth re-distribution? Wealth accumulation also invites larger philosophical questions of how society values or devalues certain activities over others.

Given these considerations, in this report, we investigate the potential sources of billionaire wealth. Specifically, we assess the question of how do wealth category, citizenship, and wealth type relate to billionaire net worths in 2014?

We hypothesize that financial wealth, European citizenship, and inheritance are related to billionaires having higher net worths. To begin with, our team hypothesizes that financial wealth relates to billionaires with higher net worths given that banking and finance have served the largest primary industries for up to 20.4% of the total billionaire population, according to the Billionaire Census 2022 (Goldberg 2022). Primary industries suggest where the billionaires devoted most of their time and by which they created most of their wealth. Second, we hypothesize that inheritance also associates with higher net worths of billionaires due to the link between increased inheritance flows and increased wealth distribution identified by several studies (Elinder 2018). Finally, we associate European citizenship with higher net worths of billionaires as the third component of the hypothesis, as Europe's billionaire wealth has been growing the fastest, up 22% year-over-year in 2021 (Avery 2022).

Our data comes from the "billionaires" data set, which we obtained from the CORGIS Dataset Project. This data set was created by Ryan Whitcomb in 2016 and originally sourced from a report by Caroline Freund and Sarah Oliver of the Peterson Institute for International Economics. Freund and Oliver derived their data set from Forbes World's Billionaires lists from 1996-2014 \[3\]. Forbes journalists drew from shareholder information, company financial statements, current exchange rates, and meetings with candidates to estimate billionaire net worths. From 1996-2010, the Forbes data included name, rank, citizenship, and net worth. Then, in 2011, Forbes added the source of wealth (company or sector).

Freund and Oliver supplemented the Forbes data with additional variables that describe the characteristics of billionaires and their businesses. The "billionaires" data set includes information on billionaires taken at three-time points: 1996, 2001, and 2014.

This report focuses on four variables: wealth, wealth type, wealth category, and citizenship. Following is a brief description of each variable of interest:

1.  Wealth - Called "wealth.worth.in.billions" in the data set. A continuous numeric variable that describes the wealth in billions for each observation, where one observation is one billionaire at one time point.

2.  Wealth type - Called "wealth.type" in the data set. A nominal categorical variable that describes the type of wealth. For billionaires in 2014, there are four wealth types: founder, inherited, privatized, and self-made.

3.  Wealth category - Called "wealth.how.category" in the data set. A nominal categorical variable that describes the category of wealth. For billionaires in 2014, there are six wealth categories: energy, financial, new, non-traded, resource, and traded.

4.  Citizenship - Called "location.citizenship" in the data set. A nominal categorical variable that describes the country of citizenship of the billionaire. In this report, we group citizenship into Europe and Not Europe.

The method by which information for this data set was collected priimarily raises two ethical concerns. First, when data collection involves human subjects, it is best practice to obtain the subjects' consent. Although the Forbes journalists conducted interviews with subjects, it is unclear whether the journalists obtained the consent of all the billionaires that they collected data on. Moreover, it is unclear whether how Forbes obtained shareholder information and company financial statements. Second, this data set contains extensive personal identifying information about individuals, which may conflict for practices of de-identifying subjects and preserving subjects' anonymity.

# Literature Review

Freund, Caroline, and Sarah Oliver. "The origins of the superrich: the billionaire characteristics database."Peterson Institute for International Economics Working Paper 16-1 (2016).

**Summary:**

This article directly addresses the origins of how billionaires get their wealth and what characteristics contributed to the rise of their wealth. The article specifies the age of billionaires as they reach the milestones of becoming millionaires and onwards. It looks into what economic sectors they work in and how much those sectors contributed to their wealth and compares the trends between Europe and America. They also compare personal factors like age and family inheritance.

**How our research differs:**

The research article analyzes and brings in new information on the impact of inheritance on accumulated wealth and net worth. We will specifically interrogate how significant of a role inheritance plays in wealth-building, which in an area that this article does not explicitly address.

**Additional published credible article:**

Sanandaji, Tino. "The international mobility of billionaires." Small Business Economics 42 (2014): 329-338.

**Summary:**

This article uses Forbes Magazine's list of billionaires and additional information to study the migration of billionaires and their sources of income that lead to them building their wealth and how their wealth interacts with the economy of the countries they live in and move to or from.

This study gives us better context on billionaires and the interactions of their wealth and movements that focuses on a couple of factors outside of the previous article and our own data set which could explain some of our own variables.

# Methodology

## Housekeeping

```{r}
#| label: load-pkgs

library(tidyverse)
library(tidymodels)

```

```{r}
#| label: load-billionaires-data

billionaires <- read_csv("data/billionaires.csv")

glimpse(billionaires)
```

```{r}
#l label: coerce-valid-var-names

names(billionaires) <- make.names(names(billionaires),
                                  unique = TRUE)
```

```{r}
#l label: 2014-only

billionaires_2014 <- billionaires |>
  filter(year == "2014")
```

```{r}
#l label: mutate-citizenship

categories_vec <- unique(billionaires$location.citizenship)
categories_tib <- enframe(categories_vec)

billionaires_2014 <- billionaires_2014 |>
  mutate(citizenship_euro = if_else(location.citizenship %in% 
                   c("Switzerland", "Spain", "Germany", "Sweden",
                     "France", "Italy", "Russia", "United Kingdom",
                     "Turkey", "Netherlands", "Greece", "Ireland",
                     "Ukraine", "Czech Republic", "Denmark", "Austria",
                     "Portugal", "Liechtenstein", "Norway", "Belgium",
                     "Poland", "Finland", "Guernsey", "Romania", "Lithuania"),
                   "Europe", "Not Europe"))
```

```{r}
#| label: summary_stats

billionaires_2014 |>
  summarize(m_wealth = mean(wealth.worth),
            me_wealth = median(wealth.worth),
            sd_wealth = sd(wealth.worth),
            na.rm = T)

billionaires_2014 |>
  group_by(wealth.how.industry) |>
  summarize(m_industry = mean(wealth.worth),
            me_industry = median(wealth.worth),
            sd_industry = sd(wealth.worth),
            na.rm = T)

billionaires_2014 |>
  group_by(wealth.how.category) |>
   summarize(m_category = mean(wealth.worth),
            me_category = median(wealth.worth),
            sd_category = sd(wealth.worth),
            na.rm = T)

billionaires_2014 |>
  group_by(location.citizenship) |>
   summarize(m_citizen = mean(wealth.worth),
            me_citizen = median(wealth.worth),
            sd_citizen = sd(wealth.worth),
            na.rm = T)

billionaires_2014 |>
  group_by(wealth.how.inherited) |>
  summarize(m_inherited = mean(wealth.worth),
            me_inherited = median(wealth.worth),
            sd_inherited = sd(wealth.worth),
            na.rm = T)

```

```{r}
#| label: box-plot-citizen

billionaires_2014 |>
ggplot(
  aes(x = citizenship_euro,
      y = wealth.worth)
  ) +
  geom_boxplot() +
  labs(title = "Citizenship vs. Net Worth",
       x = "Citizenship",
       y = "Net Worth")

```

```{r}
#| label: box-plot-cat

billionaires_2014 |>
ggplot(
  aes(x = wealth.how.category,
      y = wealth.worth)
  ) +
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1)) +
  geom_boxplot() +
  labs(title = "Wealth Category vs. Net Worth",
       x = "Wealth Category",
       y = "Net Worth")

```

```{r}
#| label: box-plot-inh 

billionaires_2014 |>
ggplot(
  aes(x = wealth.how.inherited,
      y = wealth.worth)
  ) + 
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1)) +
  geom_boxplot() +
  labs(title = "Type of Inheritance vs. Net Worth",
       x = "Inheritance",
       y = "Net Worth")

```

```{r}
#| label: proportion_stats_tables

wealth.industry_prop <- billionaires_2014 |>
  count(wealth.how.industry) |>
  group_by(wealth.how.industry %in% c()) |>
  mutate(prop_how = prop.table(n)) |>
  na.omit()

wealth.category_prop <- billionaires_2014 |>
  count(wealth.how.category) |>
  group_by(wealth.how.category %in% c()) |>
    mutate(prop_how = prop.table(n)) |>
  na.omit()

citizenship_prop <- billionaires_2014 |>
  count(location.citizenship) |>
  group_by(location.citizenship %in% c()) |>
  mutate(prop_how = prop.table(n)) |>
  na.omit()

wealth.inherited_prop <- billionaires_2014 |>
  count(wealth.how.inherited) |>
  group_by(wealth.how.inherited %in% c()) |>
  mutate(prop_how = prop.table(n)) |>
  na.omit()
```

```{r}
#l label: pie-wealth-cat

cat_prop_vec <- pull(wealth.category_prop, prop_how)

cat_labs <- c("n/a", "Energy", "Financial", "New Sectors", 
              "Non-Traded", "Resource", "Traded")

pie(cat_prop_vec,
    labels = cat_labs,
    main = "Wealth Category Proportions")

```

```{r}
#l label: pie-wealth-ind

ind_prop_vec <- pull(wealth.industry_prop, prop_how)
ind_labs <- c("n/a", "Construction", "Consumer", "Diversified Financial", 
              "Energy", "Hedge Funds", "Media", "Mining & Metals", 
              "Money Management", "Non-Consumer Industrial", "Other", 
              "Private Equity", "Real Estate", "Retail/Restaurant",
              "Tech-Computer", "Tech-Medical", "Venture Capital")

pie(ind_prop_vec,
    labels = ind_labs,
    main = "Wealth Industry Proportions")
```

```{r}
#l label: pie-inherited

inh_prop_vec <- pull(wealth.inherited_prop, prop_how)
inh_labs <- c("3rd gen", "4th gen", "5th gen", "father", "not inherited",
              "spouse/widow")

pie(inh_prop_vec,
    labels = inh_labs,
    main = "Wealth Inherited Proportions")
```

To evaluate our research question and to test our hypothesis, we selected a linear regression methodology since it is a common method to model relationships between dependent variables (billionaire net worth) and one or more independent variables (financial wealth, European citizenship, and inheritance). Moreover, it allows us to quantify the strength and direction of the relationship between the independent variables and the dependent variable. If the results of the linear regression analysis show a significant relationship between financial wealth, European citizenship, and inheritance and higher net worths of billionaires, then the data would support the hypothesis.

# Results

## Single variable linear regression models

```{r}
#l label: lin-reg-wealth-category

wealth_category <- linear_reg() |>
  set_engine("lm") |>
  fit(wealth.worth ~ wealth.how.category,
      data = billionaires_2014) |>
  glance() |>
  pull(AIC)

```

$$\widehat{wealth} = 2.69 - 0.794 * energy - 0.862 * financial + 1.61 * new
+ 1.69 * non-traded + 1.29 * resource + 1.16 * traded$$

AIC = 10474.78

```{r}
#l label: lin-reg-wealth-type

wealth_type <- linear_reg() |>
  set_engine("lm") |>
  fit(wealth.worth ~ wealth.type,
      data = billionaires_2014) |>
  glance() |>
  pull(AIC)

```

$$\widehat{wealth} = 2.70 + 1.42 * founder + 1.72 * inherited + 1.31 * privatized + 0.609 * self-made$$

AIC = 10458.57

```{r}
#l label: lin-reg-citizenship

citizenship <- linear_reg() |>
  set_engine("lm") |>
  fit(wealth.worth ~ citizenship_euro,
      data = billionaires_2014) |>
  glance() |>
  pull(AIC)
```

$$\widehat{wealth} = 4.04 - 0.193 * NotEurope$$

AIC = 10477.63

The second model, lin-reg-wealth-type, is the best fit model when looking at a single variable linear regression. This model indicates the relationship between net worth and wealth type and we know it is the best fit model for our data because the AIC value is the lowest of the 3 models. Looking at how the billionaire's accumulated their wealth or what their citizenship is shows a less accurate picture of our data set.

## Two variable linear regression models

```{r}
#| label: add-wealth-type-and-citizenship

add_model_1 <- linear_reg() |>
  set_engine("lm") |>
  fit(wealth.worth ~ wealth.type + citizenship_euro,
      data = billionaires_2014) |>
  glance() |>
  pull(AIC)
```

$$\widehat{wealth} = 2.74 + 1.42 * founder + 1.71 * inherited + 1.29 * privatized + 0.61 * self-made - 0.0557 * citizenship_euro$$

AIC = 10460.54

```{r}
#| label: int-wealth-type-and-citizenship

int_model_1 <- linear_reg() |>
  set_engine("lm") |>
  fit(wealth.worth ~ wealth.type * citizenship_euro,
      data = billionaires_2014) |>
  glance() |>
  pull(AIC)
```

$$\widehat{wealth} = 4.36 - 0.0224 * founder + 0.0178 * inherited - 0.208 * privatized - 2.03 * self-made - 2.12 * citizenship-not-euro + 1.84 * founder-not-euro + 2.18 * inerited-not-euro + 1.83 * privatized-not-euro + 3.33 * self-made-not-euro $$

AIC = 10462.5

```{r}
#| label: add-wealth-type-and-category

add_model_2 <- linear_reg() |>
  set_engine("lm") |>
  fit(wealth.worth ~ wealth.type + wealth.how.category,
      data = billionaires_2014) |>
  glance() |>
  pull(AIC)
```

$$\widehat{wealth} = 1.37 + 1.36 * founder + 1.92 * inherited + 1.68 * privatized + 1.09 * self-made - 1.15 * energy + 0.838 * financial + 1.68 * new-sectors + 1.36 * non-traded-sectors + 0.869 * resource + 1.08 * traded-sectors $$

AIC = 10458.79

```{r}
#| label: int-wealth-type-and-category

int_model_2 <- linear_reg() |>
  set_engine("lm") |>
  fit(wealth.worth ~ wealth.type * wealth.how.category,
      data = billionaires_2014) |>
  glance() |>
  pull(AIC)
```

$$\widehat{wealth} = 1.98 + 0.407 * founder + 1.54 * inherited - 0.142 * privatized - 0.514 * self-made + 0.0646 * energy + 1.84 * financial + 0.582 * new-sectors + 0.872 * non-traded-sectors +2.04 * resource $$

AIC = 10472.22

For our two variable linear regression models, the additive model "wealth.type + wealth.how.category" is the best fit for our data set. It's AIC is the lowest of the other additive model, as well as the two interaction models.

It is much lower than its counterpart interaction model of "wealth.type + wealth.how.category". This indicates that it is better for us to model one independent variable does not depend on the value of any other independent variable, than it is for us to model the effect of one variable depending on another variable.

## Three variable linear regression models

```{r}
#| label: add-three-var

add_3 <- linear_reg() |>
set_engine("lm") |>
fit(wealth.worth ~ wealth.how.category + citizenship_euro + wealth.type,
  data = billionaires_2014) |>
  glance() |>
  pull(AIC)

```

AIC = 10460.77

```{r}
#| label: int-three-var

int_model_3 <- linear_reg() |>
set_engine("lm") |>
fit(wealth.worth ~ wealth.how.category * citizenship_euro * wealth.type,
    data = billionaires_2014) |>
  glance() |>
  pull(AIC)
```

AIC = 10479.69

Our three variable linear regression models demonstrate that it is a better fit for our data set to use an additive model when comparing "wealth.how.category + citizenship_euro + wealth.type". The value of the AIC of the interaction model is higher than that of the additive.


Overall the linear regression model with the lowest AIC, therefore the best fit for our data & answering our research question is the single variable linear regression which studied the relationship between wealth.worth and wealth.type of the billionaires.

# Results  (Provide only the main results from your analysis)

Throughout this project, we set out to determine how wealth category, citizenship, and wealth type relate to billionaire net worths in the year of 2014. We hypothesized that **financial** wealth, **European** citizenship, and inheritance are linked to billionaires with higher net worths. More specifically, these variables could be indicators of a billionaire possessing an overall higher net worth.

To test our hypothesis, we began by grouping our data set by billionaires possessing citizenship in a European country and those with Non-European citizenship.




To begin with, our team hypothesizes that financial wealth relates to billionaires with higher net worths given that banking and finance have served the largest primary industries for up to 20.4% of the total billionaire population, according to the Billionaire Census 2022 (Goldberg 2022). 

Second, we hypothesize that inheritance also associates with higher net worths of billionaires due to the link between increased inheritance flows and increased wealth distribution identified by several studies (Elinder 2018). 

Finally, we associate European citizenship with higher net worths of billionaires as the third component of the hypothesis, as Europe's billionaire wealth has been growing the fastest, up 22% year-over-year in 2021 (Avery 2022).



# Works Cited

1.  Joe Hasell, Max Roser, Esteban Ortiz-Ospina and Pablo Arriagada (2022) - "Poverty". Published online at OurWorldInData.org. Retrieved from: 'https://ourworldindata.org/poverty' \[Online Resource\]

2.  Myers, J. (2021, December 10). These charts show the growing gap between the world's richest and poorest. World Economic Forum. Retrieved March 29, 2023, from https://www.weforum.org/agenda/2021/12/global-income-inequality-gap-report-rich-poor/#:\~:text=2020%20marked%20the%20steepest%20increase,1%25%20to%20over%203%25.

3.  Freund, C., & Oliver, S. (2016). (working paper). The Origins of the Superrich: The Billionaire Characteristics Database. Peterson Institute for Economics.

4.  Elinder, Mikael, et al. "Inheritance and Wealth Inequality: Evidence from Population Registers." SSRN Electronic Journal, 2016, https://doi.org/10.2139/ssrn.2757949.

5.  Goldberg, Robyn. "Billionaire Census." Altrata, 1 Aug. 2022, https://altrata.com/reports/billionaire-census-2022.

6.  Avery, Koop "Mapped: The World's Billionaire Population, by Country." Visual Capitalist, 24 Sept. 2022, https://www.visualcapitalist.com/mapped-the-worlds-billionaire-population-by-country/#:\~:text=North%20America%20is%20home%20to,net%20worth%20of%20%244.45%20trillion.&text=In%20regional%20terms%2C%20Europe's%20billionaire,%2Dover%2Dyear%20in%202021.
