---
title: "Billionaire Wealth"
subtitle: "Report"
format: html
editor: visual
execute:
  echo: true
---

# Introduction and Data

To the average person, the modern billionaire is at once mysterious and completely scrutable. Few will ever come face to face with a billionaire, yet many can name the likes of Jeff Bezos and Bill Gates and a handful of facts about each. While these billionaires sit on enormous wealth, billions of people around the globe live below the poverty. Our realities of poverty amongst wealth and stark inequality thus beg the questions how are some amassing such wealth and what factors might be correlated with wealth accumulation? 

In this report, we assess data 

# Methodology
## Housekeeping
```{r}
#| label: load-pkgs
#| message: false
#| warning: false

library(tidyverse)
library(tidymodels)

```

```{r}
#| label: load-billionaires-data

billionaires <- read_csv("data/billionaires.csv")

glimpse(billionaires)
```

```{r}
#l label: coerce-valid-var-names

names(billionaires) <- make.names(names(billionaires), unique = TRUE)
```

```{r}
#l label: 2016-only

billionaires_2016 <- billionaires |>
  filter(year == "2014")
```

```{r}
#l label: mutate-citizenship

categories_vec <- unique(billionaires$location.citizenship)
categories_tib <- enframe(categories_vec)

billionaires_2016 <- billionaires_2016 |>
  mutate(citizenship_euro = if_else(location.citizenship %in% 
                   c("Switzerland", "Spain", "Germany", "Sweden", "France", "Italy", "Russia", "United Kingdom", "Turkey", "Netherlands", "Greece", "Ireland", "Ukraine", "Czech Republic", "Denmark", "Austria", "Portugal", "Liechtenstein", "Norway", "Belgium", "Poland", "Finland", "Guernsey", "Romania", "Lithuania"), "Europe", "Not Europe"))
```

# Results
## Single variable linear regression models

```{r}
#l label: lin-reg-wealth-category

wealth_category <- linear_reg() |>
  set_engine("lm") |>
  fit(wealth.worth ~ wealth.how.category, data = billionaires_2016) |>
  glance() |>
  pull(AIC)

```

$$\widehat{wealth} = 2.69 - 0.794 * energy - 0.862 * financial + 1.61 * new
+ 1.69 * non-traded + 1.29 * resource + 1.16 * traded$$

AIC = 10474.78

```{r}
#l label: lin-reg-wealth-type

wealth_type <- linear_reg() |>
  set_engine("lm") |>
  fit(wealth.worth ~ wealth.type, data = billionaires_2016) |>
  glance() |>
  pull(AIC)

```

$$\widehat{wealth} = 2.70 + 1.42 * founder + 1.72 * inherited + 1.31 * privatized + 0.609 * self-made$$
AIC = 10458.57

```{r}
#l label: lin-reg-citizenship

citizenship <- linear_reg() |>
  set_engine("lm") |>
  fit(wealth.worth ~ citizenship_euro, data = billionaires_2016) |>
  glance() |>
  pull(AIC)
```

$$\widehat{wealth} = 4.04 - 0.193 * NotEurope$$

AIC = 10477.63